Escaping RGBland: Selecting Colors for Statistical Graphics

### 2008-10-12 pkg/paper/Palettes/CSDA.Rnw
moved hcl-colors vignette from vcd to colorspace
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@42 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

A simple and very effective palette for such a display is a set of gray colors
as in the top right panel of Figure~\ref{fig:bkde}. This is often (appropriately) used
in printed papers when the journal does not offer color graphics---however, in journals
that support color graphics (or on presentation slides and in interactive usage in statistical
software packages), many users prefer to have colored displays and most often use HSV
palettes (as in the two left two panels). These palettes code the variable of interest
by varying hue in an HSV color wheel, yielding a ``rainbow'' of colors, as done by 
\citet[Figure~8]{color:Harezlak+Coull+Laird:2007}
or \citet[Figure~2]{color:Tenenhaus+Giron+Viennet:2007}. 
The palette in the upper left panel
codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
The latter works a bit better than the former, however both palettes exhibit several drawbacks.
The modes in the map look much more like ``rings'' rather than a smoothly increasing/decreasing 
density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
it hard to hold the attention of the viewer for a longer time because the large areas shaded
with saturated colors can be distracting and produce after-image effects \citep{color:Ihaka:2003}.

@@

As our final example, Figure~\ref{fig:haireye} \citep[taken from][Figure~11, left/middle]{color:Friendly:2002}
visualizes the cross-tabulation of hair end eye color of 592 students in a mosaic display.
Clearly, hair and eye color are not independent and the pattern of association is highlighted
by means of residual-based shading. Cells associated with Pearson residuals whose absolute value exceeds
(2 or) 4 are shaded (light) blue/red. This shows that there are significantly more students
with black hair and brown eyes, blond hair and blue eyes, red hair and green eyes than expected
under independence. Conversely, fewer students than expected have blond hair and brown/hazel eyes
or black hair and blue eyes. Comparing the HSV and HCL colors in Figure~\ref{fig:haireye},
it is shown again that the HCL colors are better balanced (between the red and blue colors)
and less flashy. Even the cells associated with small residuals (below 2) are a bit easier
to read when shaded in a light grey rather than white. More details and extensions to residual-based
shadings can be found in \cite{color:Zeileis+Meyer+Hornik:2007}.

@@

Implementing the different color palettes suggested in the previous section 
is extremely easy if the software environment chosen already provides an implementation
of HCL colors: from the formulas provided above the HCL coordinates for a
palette can be conveniently computed. A bit more work is required if the software
package does not yet provide an HCL implementation. In that case, additional
functionality is needed for translating HCL coordinates to the software package's 
color system which may vary between different packages, but standardized RGB (sRGB)
is often used. The typical way of coordinate conversion is to go first from HCL to
CIELUV by simply transforming the polar $H$ and $C$ coordinates back to the original
$U$ and $V$. Subsequently, CIELUV is converted to CIEXYZ which in turn is converted
to sRGB. The details of these conversions are somewhat technical and tedious (and hence
omitted here), however the conversion formulas are still straightfoward to
implement and can, for example, be found in \cite{color:Wiki+CIELUV:2007} or
\cite{color:Poynton:2000}.

The \proglang{R} system for statistical computing \citep{color:R:2008} already comes
with an open-source implementation of HCL (and other color spaces) in the package
\pkg{colorspace} \citep{color:Ihaka:2006}. The coordinate transformations mentioned
above are contained in \proglang{C} code within \pkg{colorspace} that are easy to
port to other statistical software systems. Based on \pkg{colorspace} and
the \proglang{R} graphics system \citep[see][]{color:Murrell:2006},
the package \pkg{vcd} \citep{color:Meyer+Zeileis+Hornik:2007} provides implementations
of all palettes discussed above. Qualitative palettes are provided by
\code{rainbow_hcl()} (named after the HSV-based function \code{rainbow()} in base
\proglang{R}). Sequential palettes based on a single hue are implemented in the function
\code{sequential_hcl()} while \code{heat_hcl()} offers sequential palettes based on
a range of hues. Diverging palettes can be obtained by \code{diverge_hcl()}.
Technical documentation along with a large collection of examples is available via
\verb/help("rainbow_hcl", package = "vcd")/. Furthermore, \proglang{R} code for reproducing
Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
(and some illustrations) can be accessed via \code{vignette("hcl-colors", package = "vcd")}.

@@

We provide paths through HCL space along perceptual axes of the human visual system
so that colors selected along these paths match human perceptual dimensions.
This gives the users the possibility to conveniently experiment with the HCL-based
palettes by varying several simple and intuitive 
graphical parameters. For qualitative
palettes, these are the coordinates on the chroma and luminance axis, respectively,
controlling whether the colors are light or dark and how colorful they are.
For sequential and diverging palettes, the user can decide whether contrasts
in the chroma or luminance direction (or both) should be employed. In our
experience (as illustrated in Section~\ref{sec:motivation}), chroma contrasts
work sufficiently well if a small set of colors is used. However, when a larger
set of colors is used (e.g., for heatmaps where extreme values should be identifiable)
it is much more important to have a big difference in luminance.
%% Another parameter in these 
%% palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
%% Employing a function $|i|^p$ for this, small
%% powers $p$ can be used if palettes with a lot of color should be 
%% constructed (e.g., to clearly separate regions in a classification map). On
%% the other hand, large powers $p$ will result in palettes with less color
%% (e.g., appropriate to highlight only extreme regions firmly classified in such a map).

@@

%% Referees:
%% Ross Ihaka, Thomas Lumley, Cynthia Brewer, Martin Maechler, Ken Knoblauch

>>>>

A simple and very effective palette for such a display is a set of gray colors
as in the top right panel of Figure~\ref{fig:bkde}. This is often (appropriately) used
in printed papers when the journal does not offer color graphics---however, in journals
that support color graphics (or on presentation slides and in interactive usage in statistical
software packages), many users prefer to have colored displays and most often use HSV
palettes (as in the two left two panels). These palettes code the variable of interest
by varying hue in an HSV color wheel, yielding a ``rainbow'' of colors, as done by 
\citet[Figure~8]{color:Harezlak+Coull+Laird:2007}
or \citet[Figure~2]{color:Tenenhaus+Giron+Viennet:2007}. 
The palette in the upper left panel
codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
The latter works somewhat better than the former, however both palettes exhibit several drawbacks.
The modes in the map look much more like ``rings'' rather than a smoothly increasing/decreasing 
density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
it hard to hold the attention of the viewer for a longer time because the large areas shaded
with saturated colors can be distracting and produce after-image effects \citep{color:Ihaka:2003}.

@@

As our final example, Figure~\ref{fig:haireye} \citep[taken from][Figure~11, left/middle]{color:Friendly:2002}
visualizes the cross-tabulation of hair end eye color of 592 students in a mosaic display.
Clearly, hair and eye color are not independent and the pattern of association is highlighted
by means of residual-based shading. Cells associated with Pearson residuals whose absolute value exceeds
(2 or) 4 are shaded (light) blue/red. This shows that there are significantly more students
with black hair and brown eyes, blond hair and blue eyes, red hair and green eyes than expected
under independence. Conversely, fewer students than expected have blond hair and brown/hazel eyes
or black hair and blue eyes. Comparing the HSV and HCL colors in Figure~\ref{fig:haireye},
it is shown again that the HCL colors are better balanced (between the red and blue colors)
and less flashy. Even the cells associated with small residuals (below 2) are somewhat easier
to read when shaded in a light grey rather than white. More details and extensions to residual-based
shadings can be found in \cite{color:Zeileis+Meyer+Hornik:2007}.

@@

Implementing the different color palettes suggested in the previous section 
is extremely easy if the software environment chosen already provides an implementation
of HCL colors: from the formulas provided above the HCL coordinates for a
palette can be conveniently computed. Somewhat more work is required if the software
package does not yet provide an HCL implementation. In that case, additional
functionality is needed for translating HCL coordinates to the software package's 
color system which may vary between different packages, but standardized RGB (sRGB)
is often used. The typical way of coordinate conversion is to go first from HCL to
CIELUV by simply transforming the polar $H$ and $C$ coordinates back to the original
$U$ and $V$. Subsequently, CIELUV is converted to CIEXYZ which in turn is converted
to sRGB. The details of these conversions are somewhat technical and tedious (and hence
omitted here), however the conversion formulas are still straightfoward to
implement and can, for example, be found in \cite{color:Wiki+CIELUV:2007} or
\cite{color:Poynton:2000}.

The \proglang{R} system for statistical computing \citep{color:R:2008} provides
an open-source implementation of HCL (and other color spaces) in the package
\pkg{colorspace}, originally written by Ross Ihaka. The coordinate transformations mentioned
above are contained in \proglang{C}~code within \pkg{colorspace} that are easy to
port to other statistical software systems. Starting from version~0.96, \pkg{colorspace}
\citep{color:Ihaka+Murrell+Hornik+Zeileis:2008} also includes
an implementation of all palettes discussed above. 
\citep[Originally, the code for the palettes was in the \pkg{vcd} package,]%
[but it was recently moved to \pkg{colorspace} to be more easily acessible.]%
{color:Meyer+Zeileis+Hornik:2008} Qualitative palettes are provided by
\code{rainbow_hcl()} (named after the HSV-based function \code{rainbow()} in base
\proglang{R}). Sequential palettes based on a single hue are implemented in the function
\code{sequential_hcl()} while \code{heat_hcl()} offers sequential palettes based on
a range of hues. Diverging palettes can be obtained by \code{diverge_hcl()}.
Technical documentation along with a large collection of example palettes is available via
\verb/help("rainbow_hcl", package = "colorspace")/. Furthermore, \proglang{R}~code for reproducing
Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
(and some illustrations) can be accessed via \code{vignette("hcl-colors", package = "colorspace")}.

@@

We provide paths through HCL space along perceptual axes of the human visual system
so that colors selected along these paths match human perceptual dimensions.
This gives the users the possibility to conveniently experiment with the HCL-based
palettes by varying several simple and intuitive 
graphical parameters. For qualitative
palettes, these are the coordinates on the chroma and luminance axis, respectively,
controlling whether the colors are light or dark and how colorful they are.
For sequential and diverging palettes, the user can decide whether contrasts
in the chroma or luminance direction (or both) should be employed. In our
experience (as illustrated in Section~\ref{sec:motivation}), chroma contrasts
work sufficiently well if a small set of colors is used. However, when a larger
set of colors is used (e.g., for heatmaps where extreme values should be identifiable)
it is much more important to have a big difference in luminance.

@@
